---
title: "Research notes"
format: html
editor: visual
---

```{r setup}
library(tidyverse)
library(inequality)
library(janitor)
source(here::here("R/functions.R"))
```

## Investigating inequality indices with RECS data

As a first pass, I'm exploring what various indices look like using the RECS data. Found the very useful inequality R package.

```{r prep recs data}
#| message: false
#| warning: false
#| include: false
#| paged-print: false
recs_data <- process_2015_recs("raw-data/recs-2015") 

```

If I continue to use the inequality package, explore possible helper functions to:

- make it piping/tidy compatible by taking a data frame and then using column names

- modify the print method, to rename "income" in the output to whatever variable we used and print the within group GE value

- allow nested decomposition

### First pass – decompose theil index by income group

```{r income decomposition}
#| echo: false
#| message: true
#| warning: true
by_income <- iq_decompose(recs_data$electricity, group = as.character(recs_data$income_recs), weights = recs_data$wgt)

by_income
```

Demonstrates that virtually all the variation is within income groups and very little is between groups. This implies that total electricity consumption is not very responsive to income.

```{r inequality within income groups}
#| echo: false
#| message: true
#| warning: true

income_theil <- as_tibble(by_income$groups) |> 
  mutate(group = factor(group, 
                        levels = levels(recs_data$income_recs), 
                        ordered = T)) |> 
  rename(mean_electricity = mean_income,
         electricity_share = income_share)

income_theil |> 
  arrange(group) |> 
  ggplot(aes(x = group, y = within_ge)) + 
  geom_col() + 
  coord_flip()

```

Degree of inequality is pretty similar across income groups, with more inequality at the top and bottom of the distribution

### Second pass – decompose theil index by census division

```{r census_div decomposition}
#| echo: false 
#| message: true 
#| warning: true 

by_censusdiv <- iq_decompose(recs_data$electricity, 
                             group = as.character(recs_data$census_div), 
                             weights = recs_data$wgt)  

by_censusdiv
```

Further demonstrates minimal variation between census division and large variation between census divisions

```{r inequality within census div groups}
#| echo: false
#| message: true
#| warning: true

census_theil <- as_tibble(by_censusdiv$groups) |> 
  mutate(group = factor(group)) |> 
  rename(mean_electricity = mean_income,
         electricity_share = income_share)

census_theil |> 
  arrange(group) |> 
  ggplot(aes(x = group, y = within_ge)) + 
  geom_col() + 
  coord_flip()

```

Degree of inequality is much more variable across census divisions, with higher than usual inequality in Pacific, New England, Mid Atlantic and West/East North Central.
